---
title: "Storm Hazard Analysis"
author: "Dennis van den Berg"
date: "22/12/2015"
output: html_document
---


# Synopsis

We performed an exploratory analysis of data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in order to address the following questions: Across the United States, a) which types of events are most harmful with respect to population health in terms of fatalities and injuries, and b) which types of events have the greatest economic consequences in terms of property and crop damage? We looked at absolute effects summed per event type across all of the United States in [time period .. to ..].
[results..] Determined the top 10 event categories measured in terms of fatalities, injuries, property damage, crop damage
[discussion/in-depth] Investigate average/median damage per event type and its variability (boxplot!). Investigate number of events withing event types (boxplot per year -> shows variability)


# Data Processing

## Load libraries

```{r}
library(lattice)
library(ggplot2)
library(dplyr)
library(lubridate)
library(stringdist)
```


## Load data

```{r, cache=TRUE}
### Retrieve data set and file with event types
setwd("~/git/coursera_reproducible_research/projects/storm_hazards")
if(!file.exists("stormdata.csv.bz2")){
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv.bz2", method="curl")
}
df.stormdata <- read.csv("stormdata.csv.bz2")
df.eventtypes <- read.csv("event_types.txt")
```


## Data cleaning & selecting relevant subset


```{r}

# Convert BGN_DATE to POSIXct type
df.stormdata$BGN_DATE <- mdy_hms(df.stormdata$BGN_DATE)

# Select relevant fields: BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
df.stormdata.subset.alltime <- subset(df.stormdata, select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```


## Check data quality

```{r}
summary(df.stormdata.subset.alltime)

# Check BGN_DATE
hist(year(df.stormdata.subset.alltime$BGN_DATE))

# Check EVTYPE categories matching official categories
length(unique(df.stormdata.subset.alltime$EVTYPE))
head(sort(unique(df.stormdata.subset.alltime$EVTYPE)))
# => Many more categories (985) than expected because of non-consistent spelling

# Check NAs
sum(is.na(df.stormdata.subset.alltime))
# => No NA values

# Check PROPDMGEXP and CROPDMGEXP consistent values

sort(unique(df.stormdata.subset.alltime$PROPDMGEXP))
df.stormdata.subset.alltime[df.stormdata.subset.alltime$PROPDMGEXP %in% c("-", "?", "+", "h", "H", "0", "1", "2"),]
# => Strange values for PROPDMGEXP, all between 1993 and 1995
sort(unique(df.stormdata.subset.alltime$CROPDMGEXP))
df.stormdata.subset.alltime[df.stormdata.subset.alltime$CROPDMGEXP %in% c("?", "0", "2"),]
dim(df.stormdata.subset.alltime[df.stormdata.subset.alltime$CROPDMGEXP %in% c("?", "0", "2"),])
# => 27 strange values for CROPDMGEXP, all between 1993 and 1995

# Check for potential outliers
df.stormdata.subset.alltime[df.stormdata.subset.alltime$PROPDMGEXP %in% c("B"),]
# => Is this entry correct?
# 605953   1/1/2006 0:00:00                      FLOOD          0        0  115.00          B   32.50          M
df.stormdata.subset.alltime[df.stormdata.subset.alltime$CROPDMGEXP %in% c("B"),]
```


## Select data subset

```{r}

# Select subset in specific date range (after 1995 because of unreliable data, before 2011 because 2011 data is incomplete)
#df.stormdata.subset <- subset(df.stormdata, BGN_DATE > ymd("1995-12-31") & BGN_DATE < ymd("2006-01-01"), select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
df.stormdata.subset <- subset(df.stormdata, BGN_DATE > ymd("1995-12-31") & BGN_DATE < ymd("2011-01-01"), select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

# summary(df.stormdata.subset)
# 
# # Check BGN_DATE
# hist(year(df.stormdata.subset$BGN_DATE))
# 
# # Check EVTYPE categories matching official categories
# sort(unique(df.stormdata.subset$EVTYPE))
# length(unique(df.stormdata.subset$EVTYPE))
# 
# # Check NAs
# sum(is.na(df.stormdata.subset))
# # => No NA values
# 
# # Check PROPDMGEXP and CROPDMGEXP consistent values and potential outliers
# sort(unique(df.stormdata.subset$PROPDMGEXP))
# df.stormdata.subset[df.stormdata.subset$PROPDMGEXP %in% c("-", "?", "+", "h", "H", "0", "1", "2"),]
# summary(df.stormdata.subset[df.stormdata.subset$PROPDMGEXP=="", "PROPDMG"])
# df.stormdata.subset[df.stormdata.subset$PROPDMGEXP %in% c("B"),]
# sort(unique(df.stormdata.subset$CROPDMGEXP))
# df.stormdata.subset[df.stormdata.subset$CROPDMGEXP %in% c("?", "0", "2"),]
# dim(df.stormdata.subset[df.stormdata.subset$CROPDMGEXP %in% c("?", "0", "2"),])
# summary(df.stormdata.subset[df.stormdata.subset$CROPDMGEXP=="", "CROPDMG"])
# df.stormdata.subset[df.stormdata.subset$CROPDMGEXP %in% c("B"),]

# Replace PROPDMG * 10^PROPDMGEXP by PROPERTY_DAMAGE and CROPDMG* 10^CROPDMGEXP by CROP_DAMAGE variables
multiplier_mapping <- c("0" = 1, "1" = 10, "2" = 100, "3" = 1000, "4" = 10000, "5" = 100000, "6" = 1000000, "7" = 10000000, "8" = 100000000, "H" = 100, "h" = 100, "K" = 1000, "k" = 1000, "M" = 1000000, "m" = 1000000, "B" =1000000000, "b" = 1000000000)
df.stormdata.subset$PROPERTY_DAMAGE <- df.stormdata.subset$PROPDMG * multiplier_mapping[droplevels(df.stormdata.subset$PROPDMGEXP)]
df.stormdata.subset$CROP_DAMAGE <- df.stormdata.subset$CROPDMG * multiplier_mapping[droplevels(df.stormdata.subset$CROPDMGEXP)]
df.stormdata.subset <- subset(df.stormdata.subset, select=-c(PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```


## Clean up EVTYPE levels

```{r}

# Convert to lowercase to prevent upper/lowercase mismatch
df.stormdata.subset$EVTYPE <- tolower(df.stormdata.subset$EVTYPE)
# Manual rules for EVTYPEs with many injuries
df.stormdata.subset$EVTYPE <- gsub("tstm", "thunderstorm", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^wild/forest fire$", "wildfire", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^thunderstorm wind/hail$", "hail", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^winter weather/mix$", "winter weather", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^extreme cold$", "extreme cold/wind chill", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^snow$", "heavy snow", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^wind$", "strong wind", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^storm surge$", "storm surge/tide", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^landslide$", "debris flow", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^heavy surf/high surf$", "high surf", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^extreme windchill$", "extreme cold/wind chill", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^hurricane$", "hurricane (typhoon)", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("^record warmth$|^recored heat$", "excessive heat", df.stormdata.subset$EVTYPE)
df.stormdata.subset$EVTYPE <- gsub("wintry", "winter", df.stormdata.subset$EVTYPE)

# Try to match raw EVTYPE to correct EVENT_TYPE and add as new column to df.stormdata.subset
df.stormdata.subset$EVENT_TYPE <- df.eventtypes$EVENT_TYPE[amatch(tolower(df.stormdata.subset$EVTYPE), tolower(df.eventtypes$EVENT_TYPE), maxDist=3, nomatch=NA)]
# Use "OTHER" factor level for unmatched EVENT_TYPE
levels(df.stormdata.subset$EVENT_TYPE) <- c(levels(df.stormdata.subset$EVENT_TYPE), "OTHER")
df.stormdata.subset$EVENT_TYPE[is.na(df.stormdata.subset$EVENT_TYPE)] <- "OTHER"

# EVENT_TYPE "OTHER" is less than 2% of all entries
sum(df.stormdata.subset$EVENT_TYPE=="OTHER")/length(df.stormdata.subset$EVENT_TYPE)

# For unmatched EVENT_TYPE, list by raw EVTYPE and take top sorted by number of INJURIES
aggregate(INJURIES ~ EVTYPE, data=df.stormdata.subset[df.stormdata.subset$EVENT_TYPE=="OTHER",], FUN=length) %>% arrange(desc(INJURIES)) %>% head(10)

# Drop EVTYPE column
df.stormdata.subset <- subset(df.stormdata.subset, select=-c(EVTYPE))
```


# Results

## 1. Across the United States, which types of events are most harmful with respect to population health ("FATALITIES", "INJURIES")?

```{r}
df.fatalities <- aggregate(FATALITIES ~ EVENT_TYPE, data = df.stormdata.subset, sum)
df.injuries <- aggregate(INJURIES ~ EVENT_TYPE, data = df.stormdata.subset, sum)
df.fatalities.injuries <- merge(df.fatalities, df.injuries) %>% mutate(SUM.FATALATIES.INJURIES = FATALITIES + INJURIES)

# Top 10 event types by fatalities and injuries respectively
df.fatalities.injuries %>% arrange(desc(FATALITIES)) %>% head(10)
df.fatalities.injuries %>% arrange(desc(INJURIES)) %>% head(10)

# Plot fatalities and injuries in barplot for top categories
df.fatalities.injuries.top.sum <- df.fatalities.injuries %>% arrange(desc(SUM.FATALATIES.INJURIES)) %>% head(10)
barchart(FATALITIES + INJURIES ~ EVENT_TYPE, data=df.fatalities.injuries.top.sum, ylab="Number of people", main="Population health consequences per type")

#ggplot(df.fatalities.injuries.top.sum, aes(x = EVENT_TYPE, y = FATALITIES)) + geom_bar(stat = "identity", position = "stack")
```


## 2. Across the United States, which types of events have the greatest economic consequences?

```{r}
df.property.damage <- aggregate(PROPERTY_DAMAGE ~ EVENT_TYPE, data = df.stormdata.subset, sum)
df.crop.damage <- aggregate(CROP_DAMAGE ~ EVENT_TYPE, data = df.stormdata.subset, sum)
df.economic.damage <- merge(df.property.damage, df.crop.damage) %>% mutate(TOTAL.DAMAGE = PROPERTY_DAMAGE + CROP_DAMAGE)

# Plot damage in barplot for top categories
df.economic.damage.top.sum <- df.economic.damage %>% arrange(desc(TOTAL.DAMAGE)) %>% head(10)
barchart(PROPERTY_DAMAGE + CROP_DAMAGE ~ EVENT_TYPE, data=df.economic.damage.top.sum, ylab="Damage", main="Economic consequences per type")
```
