---
title: "Storm Hazard Analysis"
author: "Dennis van den Berg"
date: "22/12/2015"
output: html_document
---


# Synopsis

We performed an exploratory analysis of data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database in order to address the following questions: Across the United States, a) which types of events are most harmful with respect to population health in terms of fatalities and injuries, and b) which types of events have the greatest economic consequences in terms of property and crop damage? We looked at absolute effects summed per event type across all of the United States in [time period .. to ..].
[results..] Determined the top 10 event categories measured in terms of fatalities, injuries, property damage, crop damage
[discussion/in-depth] Investigate average/median damage per event type and its variability (boxplot!). Investigate number of events withing event types (boxplot per year -> shows variability)

Excessive Heat, Tornadoes, Flash Floods and Lightning are most fatal and Tornadoes, Floods, Excessive Heat and Thunderstorm Winds cause most injuries.

Thunderstorm Winds, Flash Floods, Tornadoes and Hail cause most economic damage, that damage to property is overall much higher than to crops and that the biggest contributor to crop damage is Hail.


# Data Processing

## Loading libraries and data

For our analysis we need to load the following libraries.
```{r}
library(lattice)
library(ggplot2)
library(lubridate)
library(stringdist)
suppressMessages(library(dplyr))
```

Data was downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 and we use this as our raw data set.
```{r, cache=TRUE}
if(!file.exists("stormdata.csv.bz2")){
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv.bz2", method="curl")
}
df.stormdata <- read.csv("stormdata.csv.bz2")
```

The raw data set consists of `r dim(df.stormdata)[1]` entries of `r dim(df.stormdata)[2]` variables.
```{r}
dim(df.stormdata)
names(df.stormdata)
```

Furthermore we load an the official list of event types, extracted from the National Weather Service Storm Data [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). Its contents are show below:
```{r}
eventtypes <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Designator", "Event Name", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
```
```{r}
length(eventtypes)
```


## Convert dates and select relevant variables

We first convert date variable BGN_DATE to type POSIXct
```{r}
df.stormdata$BGN_DATE <- mdy_hms(df.stormdata$BGN_DATE)
```

and select only the variables relevant for this analysis
```{r}
df.stormdata.explore <- subset(df.stormdata, select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```


## Checking data quality

A quick summary of this subsetted data set shows that there are no missing values, that a 75% majority of the dataset was generated since 1995 (see `BGN_DATE`), that multiple of the `EVTYPE` values do not correspond exactly with elements of `eventtypes`. Also, the majority of the entries seem to have zero values for `FATALITIES`, `INFURIES`, `PROPDMG` and `CROPDMG`.
```{r}
summary(df.stormdata.explore)
```

**Problem 1**: The `PROPDMGEXP` variable, which according to the documentation seems to denote exponential multiplication factors `10^{PROPDMGEXP}` for `PROPDMG` seems to contain invalid values. Also, a single amount should not be stored in two separate variables. The same goes for `CROPDMGEXP`:

```{r}
sort(unique(df.stormdata.explore$PROPDMGEXP))
sort(unique(df.stormdata.explore$CROPDMGEXP))
```

Investigating this a bit further we see that almost all entries with peculiar values for `PROPDMGEXP` and `CROPDMGEXP` are from the time range 1993 to 1995.

```{r}
summary(df.stormdata.explore[df.stormdata.explore$PROPDMGEXP %in% c("-", "?", "+", "h", "H", "0", "1", "2"),]$BGN_DATE)
summary(df.stormdata.explore[df.stormdata.explore$CROPDMGEXP %in% c("?", "0", "2"),]$BGN_DATE)
```

**Problem 2**: Furthermore we find a long list of `r length(unique(df.stormdata.explore$EVTYPE))` different `EVTYPE` event types (many more than the `r length(eventtypes)` different categories stored in our `eventtypes` variable).
```{r}
# Check EVTYPE categories matching official categories
length(unique(df.stormdata.explore$EVTYPE))
head(sort(unique(df.stormdata.explore$EVTYPE)), 10)
```


## Select date range with good data quality

We decided to select only data from 1995 until 2010 to work with, for multiple reasons: lower quality of `PROPDMGEXP` and `CROPDMGEXP` between 1993 and 1995, the fact that less event types were recorded before 1996 according to the documentation, and because the data from 2011 is not yet complete (in other words: only fully available years with higher data quality were selected).

```{r}
df.stormdata.clean <- subset(df.stormdata, BGN_DATE > ymd("1995-12-31") & BGN_DATE < ymd("2011-01-01"), select=c(BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```


## Convert property and crop damage amounts

We introduce new variables `PROPERTY_DAMAGE` and `CROP_DAMAGE` (calculated from `PROPDMG * 10^PROPDMGEXP` and `CROPDMG * 10^CROPDMGEXP` respectively) and drop the old ones.
```{r}
multiplier_mapping <- c("0" = 1, "1" = 10, "2" = 100, "3" = 1000, "4" = 10000, "5" = 100000, "6" = 1000000, "7" = 10000000, "8" = 100000000, "H" = 100, "h" = 100, "K" = 1000, "k" = 1000, "M" = 1000000, "m" = 1000000, "B" =1000000000, "b" = 1000000000)
df.stormdata.clean$PROPERTY_DAMAGE <- df.stormdata.clean$PROPDMG * multiplier_mapping[droplevels(df.stormdata.clean$PROPDMGEXP)]
df.stormdata.clean$CROP_DAMAGE <- df.stormdata.clean$CROPDMG * multiplier_mapping[droplevels(df.stormdata.clean$CROPDMGEXP)]
df.stormdata.clean <- subset(df.stormdata.clean, select=-c(PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```


## Clean up event types

We try to clean up the `EVTYPE` values by first converting them to lowercase (to prevent upper/lower mismatch) and adding a few manual substitution rules:
```{r}
df.stormdata.clean$EVTYPE <- tolower(df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("tstm", "thunderstorm", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^wild/forest fire$", "wildfire", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^thunderstorm wind/hail$", "hail", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^winter weather/mix$", "winter weather", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^extreme cold$", "extreme cold/wind chill", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^snow$", "heavy snow", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^wind$", "strong wind", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^storm surge$", "storm surge/tide", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^landslide$", "debris flow", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^heavy surf/high surf$", "high surf", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^extreme windchill$", "extreme cold/wind chill", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^hurricane$", "hurricane (typhoon)", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("^record warmth$|^recored heat$", "excessive heat", df.stormdata.clean$EVTYPE)
df.stormdata.clean$EVTYPE <- gsub("wintry", "winter", df.stormdata.clean$EVTYPE)
```

And subsequently trying to matching the raw `EVTYPE` to the officially used `eventtypes` that we defined above (correcting for small string differences). We use factor level `UNKNOWN` for an unmatched event type.

```{r}
df.stormdata.clean$EVENT_TYPE <- eventtypes[amatch(tolower(df.stormdata.clean$EVTYPE), tolower(eventtypes), maxDist=3, nomatch=NA)]
levels(df.stormdata.clean$EVENT_TYPE) <- c(levels(df.stormdata.clean$EVENT_TYPE), "UNKNOWN")
df.stormdata.clean$EVENT_TYPE[is.na(df.stormdata.clean$EVENT_TYPE)] <- "UNKNOWN"
df.stormdata.clean <- subset(df.stormdata.clean, select=-c(EVTYPE))
```

This results in `r round(sum(df.stormdata.clean$EVENT_TYPE=="UNKNOWN")/length(df.stormdata.clean$EVENT_TYPE) * 100, 2)`% of all entries having event type `UNKNOWN`, which we certainly consider acceptable.
```{r}
sum(df.stormdata.clean$EVENT_TYPE=="UNKNOWN")/length(df.stormdata.clean$EVENT_TYPE)
```


# Results

## Harmful weather events for population health

Aggregating the number of fatalities and injuries per event type:
```{r}
df.fatalities <- aggregate(FATALITIES ~ EVENT_TYPE, data = df.stormdata.clean, sum)
df.injuries <- aggregate(INJURIES ~ EVENT_TYPE, data = df.stormdata.clean, sum)
df.fatalities.injuries <- merge(df.fatalities, df.injuries) %>% mutate(SUM.FATALATIES.INJURIES = FATALITIES + INJURIES)
```

We find a top 10 event types in terms of fatalities:
```{r}
df.fatalities.injuries %>% arrange(desc(FATALITIES)) %>% head(10)
```

And a top 10 event types causing most injuries:
```{r}
df.fatalities.injuries %>% arrange(desc(INJURIES)) %>% head(10)
```

Visualizing these results shows clearly that Excessive Heat, Tornadoes, Flash Floods and Lightning are most fatal and Tornadoes, Floods, Excessive Heat and Thunderstorm Winds cause most injuries.
```{r, fig.width=10}
df.fatalities.injuries.top.sum <- df.fatalities.injuries %>% arrange(desc(SUM.FATALATIES.INJURIES)) %>% head(10)
barchart(FATALITIES + INJURIES ~ EVENT_TYPE, data=df.fatalities.injuries.top.sum, ylab="Number of people", main="Fatalities and Injuries by Weather Type")
```


## Economic consequences of bad weather events

Aggregating the amount of property and crop damage per event type:
```{r}
df.property.damage <- aggregate(PROPERTY_DAMAGE ~ EVENT_TYPE, data = df.stormdata.clean, sum)
df.crop.damage <- aggregate(CROP_DAMAGE ~ EVENT_TYPE, data = df.stormdata.clean, sum)
df.economic.damage <- merge(df.property.damage, df.crop.damage) %>% mutate(TOTAL.DAMAGE = PROPERTY_DAMAGE + CROP_DAMAGE)
```

We find a top 10 event types in terms of total damage:
```{r}
df.economic.damage %>% arrange(desc(TOTAL.DAMAGE)) %>% head(10)
```

Visualizing these results we see that Thunderstorm Winds, Flash Floods, Tornadoes and Hail cause most economic damage, that damage to property is overall much higher than to crops and that the biggest contributor to crop damage is Hail.

```{r, fig.width=10}
df.economic.damage.top.sum <- df.economic.damage %>% arrange(desc(TOTAL.DAMAGE)) %>% head(10)
barchart(PROPERTY_DAMAGE + CROP_DAMAGE ~ EVENT_TYPE, data=df.economic.damage.top.sum, ylab="Damage (USD)", main="Economic Damage by Weather Type")
```
